
\section{Nondeterministic Finite Automaton (NFA)}
% ==================================================

In automata theory, a finite state machine is called a deterministic finite automaton
(DFA), if
\begin{compactitem}
\item each of its transitions is uniquely determined by its source state and input symbol,
    and
\item reading an input symbol is required for each state transition.
\end{compactitem}

A nondeterministic finite automaton (NFA), or nondeterministic finite state machine,
needn't obey these restrictions. In particular, every DFA is also an NFA.
(via \href{http://en.wikipedia.org/wiki/Nondeterministic\_finite\_automaton}{WikiPedia})

\begin{example}[How many states do you need?]
    Given
    \[
        L = \set{ w \in \set{0,1}^* \mid \text{the $5$th digit of $w$ is $0$} }
    \]
    the DFA is easy to draw:
    \centgraph{mp/nfa-0.pdf}

    What if the digits are counted from the right, i.e.\ the last 5th digit?

    This is where NFAs could be useful. Unlike the earlier language which only takes seven
    states to create a DFA that recognize it, this one could use numbers of states since
    we don't really know how many digits to expect before reaching the one that needs to
    be $0$. An NFA on the other hand, could finish the job with merely six states:
    \centgraph{mp/nfa-1.pdf}
\end{example}

\begin{definition}[NFA]
    An NFA is a $5$-tuple
    \[ \NFA \]
    where
    \begin{compactitem}
    \item $Q$ and $\Sigma$ are finite sets
    \item $s \in Q, F \subseteq Q$
    \item $\delta \colon
        Q \times \Sigma_\varepsilon
        \footnote{$\Sigma_\varepsilon = \Sigma \cup \set\varepsilon$}
        \mapsto \powerset(Q)$
    \end{compactitem}
\end{definition}

As said, since a DFA \emph{is} an NFA, the definition of NFA is simply a generalized
version of DFA's. The difference that in an NFA a state can transit to multiple states
shows a different transition function $\delta$---the transition function of an NFA maps to
a set of states instead of exactly one state as of a DFA.

\subsection{Configurations}

Again the configurations of an NFA are very similar to that of a DFA---except that an NFA
has an $\varepsilon$ transition, a transition that is performed without requiring any
input.

\begin{definition}[Configurations of NFA]
    \begin{align*}
        \Conf &= Q \times \Sigma^*
        \\
        I(w)  &= (s, w)
        \\
        H     &= Q \times \set\varepsilon
        \\
        O(q,\varepsilon) &=
                            \begin{cases}
                                \true,  & q \in F  \\
                                \false, & \text{else}
                            \end{cases}
        \\
        R &= \set{
            (q,aw) \mapsto (q',w) \mid
            \forall q \in Q, a \in \Sigma_\varepsilon, w \in \Sigma^*
        }
    \end{align*}
\end{definition}

\begin{definition}[Language of NFA]
    \[
        L(N) = \set{ w \mid \exists \text{ accepting computation on input $w$} }
    \]
\end{definition}

\begin{example}[A simple NFA]

    The following graph shows an automaton $M$ with a binary alphabet that determines if
    the input ends with a $1$.
    \centgraph[4cm]{mp/nfa-2.pdf}

    The automaton $M$ shown above is \emph{not} a DFA since reading a $1$ in state $A$ can
    lead to $A$ or to $B$.

    $M$ in formal notation is
    \[
        M = \set{ \set{A,B}, \set{0,1}, \delta, A, \set{B} }
    \]
    where the transition function $\delta$ can be defined by the state transition table:
    \begin{center} \begin{tabular}{McMcMc}
        \hline
        ~ & 0           & 1             \\
        \hline
        A & \set{A}     & \set{A,B}     \\
        B & \varnothing & \varnothing   \\
        \hline
    \end{tabular} \end{center}
    Note that $\delta(p,1)$ has more than one state---again therefore $M$ is
    nondeterministic.

    Some possible state sequences for the input word ``1011'' are:
    \begin{center} \begin{tabular}{ l *9{Mc} }
        \hline
        input               &   & 1 &   & 0 &   & 1 &   & 1 &   \\
        \hline
        State sequence $1$  & A &   & B &   & ? &   &   &   &   \\
        State sequence $2$  & A &   & A &   & A &   & B &   & ? \\
        State sequence $3$  & A &   & A &   & A &   & A &   & B \\
        \hline
    \end{tabular} \end{center}

    The word is accepted by $M$ in sequence $3$; it doesn't matter that both other
    sequences fail to do so. In contrast, the word ``10'', which the state sequences
    for are shown below, is rejected by $M$, since there is no way to reach the only
    accepting state, $B$, by reading the final ``0'' symbol or by an
    $\varepsilon$-transition.
    \begin{center} \begin{tabular}{ l *5{Mc} }
        \hline
        input               &   & 1 &   & 0 &     \\
        \hline
        State sequence $1$  & A &   & B &   & ?   \\
        State sequence $2$  & A &   & A &   & A   \\
        \hline
    \end{tabular} \end{center}

    (Example via
    \href{http://en.wikipedia.org/wiki/Nondeterministic\_finite\_automaton}{WikiPedia})
\end{example}

\subsection{NFA \& DFA}

According to \autoref{thm:NFA_to_DFA}, any NFA can be translated to a DFA. The method is
to fully expand the NFA and draw out every branch of it, which is explained with more
details in \autoref{dem:NFA_to_DFA}.

\begin{theorem}
    \label{thm:NFA_to_DFA}
    \begin{align*}
        \forall \text{\upshape NFA} \NFA,  \\
        \exists \text{\upshape DFA} M = \lst{ Q',\Sigma,\delta',s',F' }
        \text{ s.t.\ }
        L(N) = L(M)
    \end{align*}
\end{theorem}

\begin{proof}[Proof of \autoref{thm:NFA_to_DFA}]
    \label{dem:NFA_to_DFA}
    \begin{align*}
        Q' &= \powerset(Q)  \\
        F' &= \set{ A \subseteq Q \mid A \cap F \not= \varnothing }  \\
        s' &= E\big(\set{s}\big) = \set{q \in Q \mid \exists s
                                                \xrightarrow\varepsilon q_1
                                                \xrightarrow\varepsilon q_2
                                                \xrightarrow\varepsilon q_3
                                                \cdots
                                                \xrightarrow\varepsilon q
        }  \\
        \delta'(A, a) &= E\( \operatorname*\cup_{q \in A} \delta(q,a) \)
    \end{align*}
\end{proof}

\begin{example}[Convert the following NFA to DFA]
    \label{exa:NFA_to_DFA}

    \centgraph[9cm]{mp/nfa_to_dfa-0}
    
    Start with the initial state, state $1$, form a set of its $\varepsilon$-enclosure
    $I = E(\set{s}) = \set{1,2}$.
    
    For each element of set $I$, find
    $\delta(q,a)\ \text{where}\ q \in I$
    and produce a new set
    $\set{\delta(1,a),\delta(2,a)} = \set{4,5,\varnothing} = \set{4,5}$,
    Let $I_a$ be its $\varepsilon$-enclosure
    $I_a = E(\set{4,5}) = \set{4,7,5,6,2}$.

    Similar for input symbol $b$, form
    $I_b = E(\set{\delta(1,b),\delta(2,b)}) = E(\set{3}) = \set{3,8}$.

    Make $I_a$ and $I_b$ new states and find $I_{a_a}, I_{a_b}, I_{b_a},\ \text{and}\ I_{b_b}$
    respectively; Repeat the procedure until no new states are formed.

    The following table neatly traces the whole process:
    \begin{center}
    \begin{tabular}{ *4{Ml} }
        \text{States} & I   & I_a               & I_b           \\
        \hline
        1 & \set{1,2}       & \set{4,7,5,6,2}   & \set{3,8}     \\
        2 & \set{4,7,5,6,2} & \varnothing       & \set{9,3,8}   \\
        3 & \set{3,8}       & \set{9}           & \varnothing   \\
        4 & \set{9,3,8}     & \set{9}           & \varnothing   \\
        5 & \set{9}         & \varnothing       & \varnothing   \\
    \end{tabular}
    \end{center}

    With the table, we use the $\varepsilon$-enclosure of the original initial state as
    the initial state of our DFA and those contain original final states as our new final
    states. Thus the new initial state is state $1$ and final states are $2,4,5$ (state
    $2$ contains $6$ and $7$, state $4$ and $5$ have $9$).
    
    The first row shows that state $1$ reading the input $a$ produces state
    $\set{4,7,5,6,2}$, which in the table is state $2$. Therefore in the DFA, we expect
    state $1$ transited to state $2$ with an $a$-arrow.

    With all the information we need, draw out the DFA:
    \centgraph[4.5cm]{mp/nfa_to_dfa-1}
\end{example}

